Improving \textsl{Gaia} parallax precision \\ with a data-driven model of stars

### 2017-06-14 paper/paper.tex
add bovy orcid
adding citations
abstract tweak

\author{Jo Bovy}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}
\affil{Department of Astronomy and Astrophysics, University of Toronto, 50 St. George Street, Toronto, ON M5S 3H4, Canada}

\begin{abstract}\noindent % trust me
Converting a noisy parallax measurement into a posterior
belief over distance requires inference with a prior.
Usually this prior represents beliefs about the stellar density distribution of
the Milky Way.
However, multi-band photometry exists for a large fraction of the
\gaia\ \tgas\ Catalog and is incredibly informative about stellar distances.
Here we use \tmass\ colors for $1.4$ million \tgas\ stars to build a
noise-deconvolved empirical prior distribution for stars in color--magnitude
space.
This model contains no knowledge of stellar astrophysics or the
Milky Way, but is precise because it is a generative-model fit to many noisy
parallax measurements under an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm---an Empirical Bayes approximation to a full
hierarchical model of the true parallax and photometry of every star---to
construct this prior.
The prior is combined with a \tgas\ likelihood to infer a precise photometric parallax estimate
and uncertainty (and full posterior) for every star.
Our parallax estimates are more precise than the \tgas\ catalog entries by a
median factor of 1.2 (14\% are more precise by a factor $>2$) and are more
precise than previous Bayesian distance estimates that use spatial priors.
We validate our parallax inferences using members of the Milky Way
star cluster M67, which is not visible as a cluster in the \tgas\ parallax
estimates, but appears as a cluster in our posterior parallax estimates.
Our results, including a parallax posterior pdf for each of 1.4
million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

@@

Here we present a demonstration-of-concept that shows that a
data-driven model of the \gaia\ data could in principle deliver very
precise parallax measurements, without any input of the physics or numerical models of stellar
evolution, interiors, or photospheres.
That is, this is a data-driven or purely empirical \emph{photometric
  parallax} method.
A photometric parallax is a parallax (or distance) estimate based on
the photometry (colors and apparent magnitude or magnitudes) of a
star.
Fundamentally, photometric parallaxes capitalize on the fact that the
apparent magnitude of a star is a strong function of distance, and
that the absolute magnitude is a strong function of color (or surface
temperature) and evolutionary stage.

>>>>

\author[0000-0001-6855-442X]{Jo Bovy}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}
\affil{Department of Astronomy and Astrophysics, University of Toronto, 50 St. George Street, Toronto, ON M5S 3H4, Canada}

\begin{abstract}\noindent % trust me
Converting a noisy parallax measurement into a posterior
belief over distance requires inference with a prior.
Usually this prior represents beliefs about the stellar density distribution of
the Milky Way.
However, multi-band photometry exists for a large fraction of the
\gaia\ \tgas\ Catalog and is incredibly informative about stellar distances.
Here we use \tmass\ colors for $1.4$ million \tgas\ stars to build a
noise-deconvolved empirical prior distribution for stars in color--magnitude
space.
This model contains no knowledge of stellar astrophysics or the
Milky Way, but is precise because it generates accurately a large number of noisy
parallax measurements under an assumption of stationarity; that is,
it combines the information from many stars.
We use the Extreme Deconvolution (\xd) algorithm---an Empirical Bayes approximation to a full
hierarchical model of the true parallax and photometry of every star---to
construct this prior.
The prior is combined with a \tgas\ likelihood to infer a precise photometric parallax estimate
and uncertainty (and full posterior) for every star.
Our parallax estimates are more precise than the \tgas\ catalog entries by a
median factor of 1.2 (14\% are more precise by a factor $>2$) and are more
precise than previous Bayesian distance estimates that use spatial priors.
We validate our parallax inferences using members of the Milky Way
star cluster M67, which is not visible as a cluster in the \tgas\ parallax
estimates, but appears as a cluster in our posterior parallax estimates.
Our results, including a parallax posterior pdf for each of 1.4
million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

@@

Here we present a demonstration-of-concept that shows that a
data-driven model of the \gaia\ data could in principle deliver very
precise parallax measurements, without any input of the physics or numerical models of stellar
evolution, interiors, or photospheres.
That is, this is a data-driven or purely empirical \emph{photometric
  parallax} method.
A photometric parallax is a parallax (or distance) estimate based on
the photometry (colors and apparent magnitude or magnitudes) of a
star. This idea was pioneered by \cite{juric08} using the Sloan Digital Sky Survey (\citealt{sdss}).
Fundamentally, photometric parallaxes capitalize on the fact that the
apparent magnitude of a star is a strong function of distance, and
that the absolute magnitude is a strong function of color (or surface
temperature) and evolutionary stage.
